sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2022-01-06 doc/paper.Rmd
Fix opening sentence
Edit Abstract

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data are spatially referenced observations of some spatial process---they might also be collected across time. Such data are common in ecology and can reveal ecological processes and improve conservation and management through applications such as species distribution models (SDMs). However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB). We describe the basic functionality of sdmTMB and the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
    | 4. We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
---

>>>>

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data are spatially referenced observations of some underlying spatial process---possibly collected over time. Such data are common in ecology and can reveal ecological processes and improve management through applications such as species distribution models (SDMs). However, appropriate models to analyze geostatistical data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB). We describe the basic functionality of sdmTMB and the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample model validation; (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients); and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, population abundance index standardization).
    | 4. We hope that user-friendly interfaces such as sdmTMB will open this useful class of models to a wider audience of users for applications within species distribution modelling and beyond.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder
---
